Numerical Stability of Detonations in White Dwarf Simulations

### 2018-07-13 papers/ignition/paper.tex
Move burner to its own section

% Software names
\newcommand{\amrex}{\texttt{AMReX}}
\newcommand{\boxlib}{\texttt{BoxLib}}
\newcommand{\castro}{\texttt{CASTRO}}
\newcommand{\microphysics}{\texttt{Microphysics}}
\newcommand{\wdmerger}{\texttt{wdmerger}}
\newcommand{\python}{\texttt{Python}}
\newcommand{\matplotlib}{\texttt{matplotlib}}
\newcommand{\yt}{\texttt{yt}}
\newcommand{\vode}{\texttt{VODE}}
\newcommand{\isoseven}{\texttt{iso7}}
\newcommand{\aproxthirteen}{\texttt{aprox13}}
\newcommand{\aproxnineteen}{\texttt{aprox19}}
\newcommand{\aproxtwentyone}{\texttt{aprox21}}

@@

\section{Nuclear Burning Methodology}
\label{sec:methodology}

In this section we discuss several issues in reactive hydrodynamic simulations and
how we addressed them in the \castro\ code \citep{castro}. We first describe the
core operation of our burner module, which is implemented in the \microphysics\ code
repository\footnote{\microphysics\ can be obtained at \url{https://github.com/starkiller-astro/Microphysics}.},
a collection of microphysical routines that are compatible with \castro.

\subsection{Nuclear Burning}
\label{sec:burner}

Given a set of nuclides and the reaction links between them, we now consider
how a burning step is performed in our software. The goal is to integrate the
vector ${\bm{Y}} = (Y_1, Y_2, \ldots, Y_n, e, T)$, where $Y_{n} = X_{n} / A_{n}$
is the molar fraction of species $n$, with $X_n$ the mass fraction and $A_n$ the
mass number of that species, $e$ is the energy released during the burn, and
$T$ is the temperature. The equation describing its evolution is given by
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y}),
\end{equation}
where the components of the right-hand-side for the species come from the particular
nuclear burning network we are using. The energy $e$ of the zone
will change when the nuclear abundances evolve, according to
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

We define several burning modes that determine how $T$ and $e$ are evolved
during a nuclear burn. In a hydrostatic burn, which we call burning mode 0,
we keep $\rho$ and $T$ fixed throughout, and use 
the energy released at the end to compute a final temperature that is
thermodynamically consistent with the new internal energy. By contrast,
in a self-heating burn (mode 1), we allow the temperature to evolve in response
to the burning (see\footnote{In the cited paper, a term based on the
thermodynamic chemical potential was included; we now believe
that it is incorrect to include such a term in the burn, since it
automatically sums to zero analytically.} \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\frac{\partial e}{\partial t}
\end{equation}
(Although $T$ evolves during the burn so that the integration is physically
accurate, as in the hydrostatic method we discard the final value
for $T$ at the end of the burn and recompute a temperature for the zone that is
consistent with its new internal energy.) Here $c_V$ is the specific heat at
constant volume, which is provided by the equation of state.  During this burn,
we can keep $c_V$ constant using its initial value, or at each step we
can choose to re-evaluate the equation of state using the latest value of $(\rho, T)$.
The latter is more expensive but also more accurate, and we use it in this paper.
In practice we find that the cost is small in comparison to the more expensive
parts of the calculation, and it can significantly speed up convergence near NSE.
A third option (mode 2) presented by \citet{raskin:2010} is a so-called ``hybrid'' mode.
In this mode, by default we do a hydrostatic burn. If that burn fails, or if the net
energy change is negative, we do the burn again in self-heating mode. A final option (mode 3)
is a burn that limits the changes due to a burn to avoid numerically unstable burning.
This mode is discussed in \autoref{sec:unstable_burning} and we will call it a
``suppressed'' burn for the remainder of this work. All four options
are implemented in our burner software.

The simulations shown in this work all use the self-heating mode. The hydrostatic
and hybrid burns are only a slight variation on the self-heating burn in practice
for this problem.
The reason that the hydrostatic burn cannot substantially change the outcome is
that even though the temperature is fixed during the burn, it is \textit{not}
constant in the rest of the hydrodynamics step. And in the hydrodynamics step,
the \castro\ algorithm performs multiple EOS calls to ensure that the temperature
is synchronized with the internal energy at various points in the update. So in
practice there is a temperature change due to the burn for our problem. This seems
inevitable for any hydro code unless the hydrodynamics update includes an explicit
equation for $T$ that is independent of what is happening for the internal energy.
The hydrostatic burn does slightly limit the energy release during any given burn,
but summed up over the whole simulation this is not enough to qualitatively affect
the outcome. The hybrid burn does not affect the simulation outcome at all relative
to the hydrostatic burn. The difference between them only applies for zones that have
a net negative energy release from the hydrostatic burn, which occurs only after
we have burned to NSE. The hybrid burn was mainly implemented by \citet{raskin:2010}
for making it easier to numerically complete a burn, rather than because of the
effects it has on the outcome of the simulation. The suppressed burn has separate
issues which we defer to \autoref{sec:unstable_burning}.

In our \microphysics\ repository we provide several software options for
solving a set of coupled stiff ODEs. For this work our primary integrator
is the classic \vode\ integrator \citep{vode}, of which we have a version
compatible with our software interfaces in the \microphysics\ repository.
We also employ an implementation of the well known variable-order Richardson
extrapolation method presented by \citet{stoer:1980}, that is similar to the
integrator which ships with the original versions of the networks mentioned
above. Our experience shows that relying on a single ODE integrator for the
variety of stiff ODEs that are present in a collision is insufficient, as we
sometimes run into integration failures. A combination of the two integrators
is usually sufficient, however: we first try \vode, and if that fails, we then
try the Stoer and Bulirsch method. Since it is possible for neither method to
succeed, we have one more backup scheme. Our specified (relative and absolute)
error tolerances are uniformly $10^{-6}$ for the energy, temperature, and nuclear
abundance ODEs. We can loosen these tolerances in an attempt to give the integrator
an easier time. If both \vode\ and the Stoer and Bulirsch algorithm fail, then we
loosen the tolerances by a multiplicative factor of 25\%, and then repeat the attempts
with \vode\ and Stoer and Bulirsch. We keep doing this until either we have loosened
the tolerances enough for one of the two methods to succeed, or until we have loosened
the tolerances by two orders of magnitude, at which point we finally give up. We have
found that this scheme is resilient enough to handle all of the simulations shown below.
We have also verified that our results do not depend strongly on the tolerances when they
are tighter than $10^{-4}$. For the typical collision problem we discuss below, the
nucleosynthetic yield for nickel varies by less than 5\% between a tolerance of $10^{-4}$
and a tolerance of $10^{-12}$.

\subsection{Coupling to the Hydrodynamics}
\label{sec:hydrocoupling}

@@

>>>>

% Software names
\newcommand{\amrex}{\texttt{AMReX}}
\newcommand{\boxlib}{\texttt{BoxLib}}
\newcommand{\castro}{\texttt{CASTRO}}
\newcommand{\maestro}{\texttt{Maestro}}
\newcommand{\microphysics}{\texttt{Microphysics}}
\newcommand{\wdmerger}{\texttt{wdmerger}}
\newcommand{\python}{\texttt{Python}}
\newcommand{\matplotlib}{\texttt{matplotlib}}
\newcommand{\yt}{\texttt{yt}}
\newcommand{\vode}{\texttt{VODE}}
\newcommand{\isoseven}{\texttt{iso7}}
\newcommand{\aproxthirteen}{\texttt{aprox13}}
\newcommand{\aproxnineteen}{\texttt{aprox19}}
\newcommand{\aproxtwentyone}{\texttt{aprox21}}

@@

\section{Nuclear Burning Methodology}
\label{sec:burner}

We first describe the core operation of our burner module, which is implemented in
the \microphysics\ code repository\footnote{\microphysics\ can be obtained at
  \url{https://github.com/starkiller-astro/Microphysics}.}, a collection of microphysical
routines that are compatible with \castro, \maestro, and other \amrex-based astrophysics
codes. We review this because, even though the ideas are not new,
it serves as an update on the state of the \castro\ code since the original release of
the code paper, similar to the goal of \cite{wdmergerI} for the hydrodynamics and gravity
modules.

Given a set of nuclides and the reaction links between them, we consider
how a burning step is performed in our software. The goal is to integrate the
vector ${\bm{Y}} = (Y_1, Y_2, \ldots, Y_n, e, T)$, where $Y_{n} = X_{n} / A_{n}$
is the molar fraction of species $n$, with $X_n$ the mass fraction and $A_n$ the
mass number of that species, $e$ is the energy released during the burn, and
$T$ is the temperature. The equation describing its evolution is given by
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y}),
\end{equation}
where the components of the right-hand-side for the species come from the particular
nuclear burning network we are using. The energy $e$ of the zone
will change when the nuclear abundances evolve, according to
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

For most of the problems of interest to \castro\, we employ a self-heating burn,
where we allow the temperature to evolve in response
to the burning (see\footnote{In the cited paper, a term based on the
thermodynamic chemical potential was included; we now believe
that it is incorrect to include such a term in the burn, since it
automatically sums to zero analytically.} \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\frac{\partial e}{\partial t}
\end{equation}
Here $c_V$ is the specific heat at constant volume, which
is provided by the equation of state. During this burn, we re-evaluate $c_V$
by calling the equation of state using the latest value of $(\rho, T)$.
The added expense of the equation of state call (relative to keeping $c_V$ fixed
through the burn) is small in comparison to the more expensive parts of the calculation,
and it can significantly speed up convergence near NSE. Although $T$ evolves during the
burn so that the computation of the reaction rates is accurate, we discard the final value
for $T$ at the end of the burn and recompute a temperature for the zone that is consistent
with its new internal energy according to the equation of state.

In our \microphysics\ repository we provide several software options for
solving a set of coupled stiff ODEs. For this work our primary integrator
is the classic \vode\ integrator \citep{vode}, of which we have a version
compatible with our software interfaces in the \microphysics\ repository.
Another option is an implementation of the well known variable-order Richardson
extrapolation method presented by \citet{stoer:1980}.

\section{Coupling Burning and Hydrodynamics}
\label{sec:hydrocoupling}

@@

% Sticking this here for now, needs to be updated with new retry discussion

Our experience shows that relying on a single ODE integrator for the
variety of stiff ODEs that are present in a collision is insufficient, as we
sometimes run into integration failures. A combination of the two integrators
is usually sufficient, however: we first try \vode, and if that fails, we then
try the Stoer and Bulirsch method. Since it is possible for neither method to
succeed, we have one more backup scheme. Our specified (relative and absolute)
error tolerances are uniformly $10^{-6}$ for the energy, temperature, and nuclear
abundance ODEs. We can loosen these tolerances in an attempt to give the integrator
an easier time. If both \vode\ and the Stoer and Bulirsch algorithm fail, then we
loosen the tolerances by a multiplicative factor of 25\%, and then repeat the attempts
with \vode\ and Stoer and Bulirsch. We keep doing this until either we have loosened
the tolerances enough for one of the two methods to succeed, or until we have loosened
the tolerances by two orders of magnitude, at which point we finally give up. We have
found that this scheme is resilient enough to handle all of the simulations shown below.
We have also verified that our results do not depend strongly on the tolerances when they
are tighter than $10^{-4}$. For the typical collision problem we discuss below, the
nucleosynthetic yield for nickel varies by less than 5\% between a tolerance of $10^{-4}$
and a tolerance of $10^{-12}$.
